---
title: "ECON 216 EDA — Motorcycle Safety"
author: "Arya, Gayathri, Skylar"
output: pdf_document
---
```{r}
library(tidyverse)


 
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(readr)
library(lubridate)
```

This document builds the shared FARS analysis dataset used in later sections. The goal here is only to load the yearly raw files, align schema differences across years, and create a cleaned joined table that downstream summaries can reuse.

```{r load-raw-data}
years <- 2011:2020

acc_list <- vector("list", length(years))
veh_list <- vector("list", length(years))
per_list <- vector("list", length(years))

for (i in seq_along(years)) {
  year <- years[i]

  if (year == 2020) {
    base_path <- file.path("data", "FARS2020NationalCSV", "FARS2020NationalCSV")
    accident_path <- file.path(base_path, "accident.csv")
    vehicle_path <- file.path(base_path, "vehicle.csv")
    person_path <- file.path(base_path, "person.csv")
  } else {
    base_path <- file.path("data", paste0("FARS", year, "NationalCSV"))
    accident_path <- file.path(base_path, "ACCIDENT.CSV")
    vehicle_path <- file.path(base_path, "VEHICLE.CSV")
    person_path <- file.path(base_path, "PERSON.CSV")
  }

  acc_list[[i]] <- read_csv(accident_path, show_col_types = FALSE) %>%
    mutate(
      across(ends_with("NAME"), as.character),
      across(starts_with("VIN"), as.character),
      YEAR = year
    )

  veh_list[[i]] <- read_csv(vehicle_path, show_col_types = FALSE) %>%
    mutate(
      across(ends_with("NAME"), as.character),
      across(starts_with("VIN"), as.character),
      across(starts_with("TRLR"), as.character),
      YEAR = year
    )

  per_list[[i]] <- read_csv(person_path, show_col_types = FALSE) %>%
    mutate(
      across(ends_with("NAME"), as.character),
      across(starts_with("VIN"), as.character),
      YEAR = year
    )
}

acc_raw <- dplyr::bind_rows(acc_list)
veh_raw <- dplyr::bind_rows(veh_list)
per_raw <- dplyr::bind_rows(per_list)

cat("Rows in acc_raw:", nrow(acc_raw), "\n")
cat("Rows in veh_raw:", nrow(veh_raw), "\n")
cat("Rows in per_raw:", nrow(per_raw), "\n")
```

The loading step loops over 2011 through 2020, adds a numeric `YEAR` field to each source table, and uses `bind_rows()` so columns that appear only in later years are still preserved. This is important because the 2020 files have a different directory structure and more descriptive columns than earlier files.

```{r join-raw-tables}
acc_veh <- acc_raw %>%
  left_join(veh_raw, by = c("ST_CASE", "YEAR"))

fars_full <- acc_veh %>%
  left_join(per_raw, by = c("ST_CASE", "YEAR"))

cat("Rows in acc_veh:", nrow(acc_veh), "\n")
cat("Rows in fars_full before cleaning:", nrow(fars_full), "\n")
```

The joins are done on both `ST_CASE` and `YEAR` so crashes from different years are never accidentally merged together. Because one crash can involve multiple vehicles and multiple people, row counts increase after each join and that expansion is expected.

```{r clean-and-filter-data}
key_factor_columns <- c(
  "STATENAME", "WEATHERNAME", "LGT_CONDNAME", "RUR_URBNAME", "SEXNAME",
  "PER_TYPNAME", "INJ_SEVNAME", "REST_USENAME", "EJECTIONNAME",
  "DRINKINGNAME", "FUNC_SYSNAME", "MAN_COLLNAME", "DAY_WEEKNAME",
  "ROLLOVERNAME", "DEFORMEDNAME", "L_STATUSNAME"
)

fars_full <- fars_full %>%
  mutate(
    AGE = as.numeric(AGE),
    FATALS = as.numeric(FATALS),
    DRUNK_DR = as.numeric(DRUNK_DR),
    TRAV_SP = as.numeric(TRAV_SP),
    BODY_TYP = as.numeric(BODY_TYP),
    AGE = na_if(AGE, 99),
    TRAV_SP = if_else(TRAV_SP %in% c(997, 998, 999), NA_real_, TRAV_SP),
    vehicle_type = case_when(
      !is.na(BODY_TYPNAME) & str_detect(BODY_TYPNAME, regex("motorcycle", ignore_case = TRUE)) ~ "Motorcycle",
      !is.na(BODY_TYPNAME) & str_detect(
        BODY_TYPNAME,
        regex("sedan|coupe|hatchback|convertible|station wagon|4-door|2-door|passenger car", ignore_case = TRUE)
      ) ~ "Car",
      is.na(BODY_TYPNAME) & BODY_TYP %in% 80:89 ~ "Motorcycle",
      is.na(BODY_TYPNAME) & BODY_TYP %in% 1:9 ~ "Car",
      TRUE ~ NA_character_
    ),
    helmet_used = case_when(
      (!is.na(REST_USENAME) & str_detect(REST_USENAME, regex("helmet", ignore_case = TRUE))) |
        (!is.na(HELM_USENAME) & str_detect(HELM_USENAME, regex("helmet", ignore_case = TRUE))) ~ TRUE,
      is.na(REST_USENAME) & is.na(HELM_USENAME) ~ NA,
      (!is.na(REST_USENAME) | !is.na(HELM_USENAME)) ~ FALSE,
      TRUE ~ NA
    )
  ) %>%
  mutate(
    across(any_of(key_factor_columns), as.factor),
    vehicle_type = factor(vehicle_type, levels = c("Motorcycle", "Car"))
  )

# Keep only motorcycle and passenger-car records because the project compares these vehicle groups.
fars_full <- fars_full %>%
  filter(!is.na(vehicle_type))

cat("Rows in fars_full after vehicle filtering:", nrow(fars_full), "\n")
```

This cleaning step standardizes missing-value codes, creates the two-category `vehicle_type` variable despite schema drift across years, and builds a unified `helmet_used` field from the restraint and helmet name columns. The factor conversion keeps key descriptive variables ready for later grouped summaries and plots.

```{r spot-check-clean-data}
print(head(fars_full[, c("ST_CASE", "YEAR", "vehicle_type", "helmet_used", "FATALS", "AGE")], 3))
```

The final spot check confirms that the joined table contains the core identifiers and cleaned variables needed for the next EDA sections. At this stage, the document has only prepared the analysis dataset and documented the one explicit filtering step used to focus on motorcycles and passenger cars.

## Section A — Univariate Visualizations

```{r section-a-helpers, echo=TRUE}
collapse_plot_categories <- function(values, top_n = 15, prop = 0.02) {
  values <- forcats::fct_lump_prop(values, prop = prop, other_level = "Other")

  if (nlevels(values) > top_n) {
    values <- forcats::fct_lump_n(values, n = top_n - 1, other_level = "Other")
  }

  forcats::fct_infreq(values)
}

plot_categorical_share <- function(var_name, plot_title, x_label, top_n = 15, prop = 0.02) {
  plot_tbl <- fars_full %>%
    filter(!is.na(.data[[var_name]])) %>%
    mutate(plot_value = collapse_plot_categories(as.factor(.data[[var_name]]), top_n = top_n, prop = prop)) %>%
    count(plot_value) %>%
    mutate(
      pct = n / sum(n),
      plot_value = forcats::fct_reorder(plot_value, pct, .desc = FALSE)
    )

  ggplot(plot_tbl, aes(x = plot_value, y = pct)) +
    geom_col(fill = "steelblue") +
    scale_y_continuous(labels = scales::percent_format()) +
    coord_flip() +
    labs(title = plot_title, x = x_label, y = "Proportion of Records") +
    theme_minimal()
}
```

```{r hist-fatals, echo=TRUE}
fars_full %>%
  filter(!is.na(FATALS)) %>%
  ggplot(aes(x = FATALS)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", boundary = 0) +
  labs(
    title = "Distribution of Fatalities per Crash Record",
    x = "Number of Fatalities",
    y = "Count of Records"
  ) +
  theme_minimal()
```

Most records cluster at the lowest fatality counts, with frequencies dropping off quickly as fatalities increase. The shape suggests that higher-fatality crash records are comparatively uncommon in this analysis sample.

```{r hist-age, echo=TRUE}
fars_full %>%
  filter(!is.na(AGE)) %>%
  ggplot(aes(x = AGE)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", boundary = 0) +
  labs(
    title = "Distribution of Person Age",
    x = "Age (years)",
    y = "Count of Records"
  ) +
  theme_minimal()
```

The age distribution is concentrated in adult ages, with relatively few records at very young or very old ages. This pattern indicates that the joined person-level records are dominated by working-age and middle-age individuals.

```{r hist-trav-sp, echo=TRUE}
fars_full %>%
  filter(!is.na(TRAV_SP)) %>%
  ggplot(aes(x = TRAV_SP)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", boundary = 0) +
  labs(
    title = "Distribution of Reported Travel Speed",
    x = "Travel Speed (mph)",
    y = "Count of Records"
  ) +
  theme_minimal()
```

Reported travel speeds are concentrated in lower to moderate speed ranges, with a thinner right tail at higher speeds. The distribution is right-skewed, so a small share of observations accounts for the fastest reported travel speeds.

```{r hist-drunk-dr, echo=TRUE}
fars_full %>%
  filter(!is.na(DRUNK_DR)) %>%
  ggplot(aes(x = DRUNK_DR)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", boundary = 0) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "Distribution of Drunk Drivers Count",
    x = "Number of Drunk Drivers",
    y = "Count of Records"
  ) +
  theme_minimal()
```

Most observations fall at zero or one drunk driver, and larger counts appear much less frequently. This makes the variable highly discrete and concentrated at the low end of its support.

```{r bar-vehicle-type, echo=TRUE}
plot_categorical_share(
  var_name = "vehicle_type",
  plot_title = "Vehicle Type Composition in the Analysis Sample",
  x_label = "Vehicle Type",
  top_n = 15,
  prop = 0.02
)
```

The sample is split across the two project vehicle groups, so this bar chart shows the relative prevalence of motorcycles and passenger cars after filtering. It provides a quick baseline for how balanced the comparison set is before moving to grouped analysis.

```{r bar-statename, echo=TRUE}
plot_categorical_share(
  var_name = "STATENAME",
  plot_title = "Crash Records by State (Top 15 + Other)",
  x_label = "State",
  top_n = 16,
  prop = 0
)
```

The state distribution is concentrated in a limited set of high-count states, while the remaining states are grouped into Other. Because the `STATENAME` field is only available for 2020 in this pipeline, this chart should be interpreted as a 2020-only descriptive snapshot.

```{r bar-weathername, echo=TRUE}
plot_categorical_share(
  var_name = "WEATHERNAME",
  plot_title = "Crash Records by Weather Condition",
  x_label = "Weather Condition",
  top_n = 15,
  prop = 0.02
)
```

Weather categories are usually dominated by clear-condition records, with smaller shares spread across rain, snow, fog, and less common conditions. The chart summarizes how strongly the observations are concentrated in a few broad weather categories.

```{r bar-lgt-condname, echo=TRUE}
plot_categorical_share(
  var_name = "LGT_CONDNAME",
  plot_title = "Crash Records by Light Condition",
  x_label = "Light Condition",
  top_n = 15,
  prop = 0.02
)
```

Light-condition counts often cluster in daylight and dark-but-lighted settings, with smaller shares in less common visibility categories. This distribution gives context on the ambient lighting environments represented in the data.

```{r bar-rur-urbname, echo=TRUE}
plot_categorical_share(
  var_name = "RUR_URBNAME",
  plot_title = "Crash Records by Rural or Urban Setting",
  x_label = "Rural or Urban Setting",
  top_n = 15,
  prop = 0.02
)
```

The rural-urban split shows whether the usable records are concentrated in more urbanized or more rural roadway settings. This is a compact way to summarize the geographic context of the crash records that survive the filtering steps.

```{r bar-sexname, echo=TRUE}
plot_categorical_share(
  var_name = "SEXNAME",
  plot_title = "Crash Records by Person Sex",
  x_label = "Person Sex",
  top_n = 15,
  prop = 0.02
)
```

This chart shows the sex composition of the person-level records included in `fars_full`. Any imbalance here is descriptive of the observed sample and helps frame later comparisons without implying a mechanism.

```{r bar-inj-sevname, echo=TRUE}
plot_categorical_share(
  var_name = "INJ_SEVNAME",
  plot_title = "Crash Records by Injury Severity",
  x_label = "Injury Severity",
  top_n = 15,
  prop = 0.02
)
```

Injury severity categories summarize how records are distributed across fatal, incapacitating, possible, and other injury outcomes. The relative bar heights show which severity categories are most common in the joined sample.

```{r bar-rest-usename, echo=TRUE}
plot_categorical_share(
  var_name = "REST_USENAME",
  plot_title = "Crash Records by Restraint or Helmet Use Category",
  x_label = "Restraint or Helmet Use",
  top_n = 11,
  prop = 0.02
)
```

`REST_USENAME` contains many detailed categories, so the chart keeps only the most common groups visible and pools the remainder into Other. That aggregation makes the restraint-use pattern easier to read while preserving the main distributional differences.

```{r bar-ejectionname, echo=TRUE}
plot_categorical_share(
  var_name = "EJECTIONNAME",
  plot_title = "Crash Records by Ejection Status",
  x_label = "Ejection Status",
  top_n = 15,
  prop = 0.02
)
```

Ejection status is typically concentrated in a few categories, so the plot highlights whether most records involve no ejection, partial ejection, or complete ejection. The compact set of bars makes it easy to compare the dominant outcomes.

```{r bar-drinkingname, echo=TRUE}
plot_categorical_share(
  var_name = "DRINKINGNAME",
  plot_title = "Crash Records by Drinking Involvement",
  x_label = "Drinking Involvement",
  top_n = 15,
  prop = 0.02
)
```

The drinking-involvement distribution shows how records divide across reported drinking categories after missing values are removed. A strong concentration in one category would indicate that the observed sample is not evenly spread across these classifications.

```{r bar-func-sysname, echo=TRUE}
plot_categorical_share(
  var_name = "FUNC_SYSNAME",
  plot_title = "Crash Records by Functional System",
  x_label = "Functional System",
  top_n = 15,
  prop = 0.02
)
```

Functional-system categories describe the roadway class associated with each record, and the chart highlights the largest groups directly. Collapsing small categories prevents the long tail of roadway types from overwhelming the main pattern.

```{r bar-man-collname, echo=TRUE}
plot_categorical_share(
  var_name = "MAN_COLLNAME",
  plot_title = "Crash Records by Manner of Collision",
  x_label = "Manner of Collision",
  top_n = 15,
  prop = 0.02
)
```

The manner-of-collision distribution shows which crash configurations appear most often in the available records. Grouping rarer patterns into Other keeps the figure readable while still preserving the broad shape of the distribution.

```{r bar-day-weekname, echo=TRUE}
plot_categorical_share(
  var_name = "DAY_WEEKNAME",
  plot_title = "Crash Records by Day of Week",
  x_label = "Day of Week",
  top_n = 15,
  prop = 0.02
)
```

Day-of-week counts indicate whether crash records are spread fairly evenly across the week or more concentrated on specific days. This is a simple descriptive view of timing within the weekly cycle.

```{r bar-deformedname, echo=TRUE}
plot_categorical_share(
  var_name = "DEFORMEDNAME",
  plot_title = "Crash Records by Vehicle Deformation Category",
  x_label = "Vehicle Deformation Category",
  top_n = 15,
  prop = 0.02
)
```

Vehicle deformation categories summarize the reported severity of physical damage in the vehicle records. The plot shows whether observations are concentrated in a few broad damage levels or dispersed across many categories.

## Section B — Relationship Visualizations

```{r section-b-helpers, echo=TRUE}
crash_vehicle_records <- fars_full %>%
  distinct(
    ST_CASE,
    YEAR,
    vehicle_type,
    FATALS,
    STATENAME,
    WEATHERNAME,
    LGT_CONDNAME,
    RUR_URBNAME
  )

motorcycle_records <- fars_full %>%
  filter(vehicle_type == "Motorcycle")

motorcycle_crash_records <- crash_vehicle_records %>%
  filter(vehicle_type == "Motorcycle")

top_motorcycle_states <- motorcycle_records %>%
  filter(!is.na(STATENAME)) %>%
  count(STATENAME, sort = TRUE) %>%
  slice_head(n = 15) %>%
  pull(STATENAME) %>%
  as.character()

top_injury_levels <- fars_full %>%
  filter(!is.na(INJ_SEVNAME)) %>%
  count(INJ_SEVNAME, sort = TRUE) %>%
  slice_head(n = 4) %>%
  pull(INJ_SEVNAME) %>%
  as.character()
```

```{r bar-fatality-by-vehicle-type, echo=TRUE}
fatality_by_vehicle_type <- crash_vehicle_records %>%
  filter(!is.na(vehicle_type), !is.na(FATALS)) %>%
  group_by(vehicle_type) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop")

ggplot(fatality_by_vehicle_type, aes(x = vehicle_type, y = total_fatalities, fill = vehicle_type)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Total Fatalities by Vehicle Type",
    x = "Vehicle Type",
    y = "Fatality Count"
  ) +
  theme_minimal()
```

This opening comparison shows that the total fatalities attached to motorcycle-involved crash records are high even though motorcycles represent a much smaller vehicle class than passenger cars. According to IIHS (2024), motorcyclists account for a disproportionate share of traffic fatalities relative to their share of registered vehicles, so this first plot sets up that broader descriptive imbalance. The next question is whether that difference remains once the comparison is normalized into a fatal-crash rate rather than a raw count.

```{r bar-fatality-proportion-by-vehicle-type, echo=TRUE}
fatality_proportion_by_vehicle_type <- crash_vehicle_records %>%
  filter(!is.na(vehicle_type), !is.na(FATALS)) %>%
  group_by(vehicle_type) %>%
  summarise(fatality_rate = mean(FATALS > 0), .groups = "drop")

ggplot(fatality_proportion_by_vehicle_type, aes(x = vehicle_type, y = fatality_rate, fill = vehicle_type)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Share of Crash Records with at Least One Fatality by Vehicle Type",
    x = "Vehicle Type",
    y = "Proportion of Crash Records with FATALS > 0"
  ) +
  theme_minimal()
```

Normalizing the comparison makes the vehicle-type gap easier to interpret because it focuses on the share of crash records that include at least one fatality. If the motorcycle bar still sits above the car bar, that suggests the difference is not only about sample size but also about a higher concentration of fatal outcomes within motorcycle-involved crashes. That motivates looking at the full shape of the fatality distribution rather than only a binary fatal versus non-fatal split.

```{r hist-fatals-faceted, echo=TRUE}
crash_vehicle_records %>%
  filter(!is.na(FATALS), !is.na(vehicle_type)) %>%
  ggplot(aes(x = FATALS)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", boundary = 0) +
  facet_wrap(~ vehicle_type, scales = "free_y") +
  labs(
    title = "Distribution of Fatalities per Crash Record by Vehicle Type",
    x = "Number of Fatalities",
    y = "Count of Crash Records"
  ) +
  theme_minimal()
```

The faceted histogram shows whether one vehicle group has a heavier right tail in fatalities per crash record, not just a higher share of fatal crashes overall. Both panels are likely concentrated at low fatality counts, but the relative mass in the higher-fatality bins helps clarify whether motorcycle crashes are more tightly clustered near severe outcomes. After that distributional check, the analysis can move from crash-level fatalities to person-level injury severity patterns.

```{r bar-inj-sev-by-vehicle-type, echo=TRUE}
injury_severity_by_vehicle_type <- fars_full %>%
  filter(!is.na(vehicle_type), !is.na(INJ_SEVNAME)) %>%
  count(vehicle_type, INJ_SEVNAME) %>%
  group_by(vehicle_type) %>%
  mutate(injury_share = n / sum(n)) %>%
  ungroup() %>%
  mutate(INJ_SEVNAME = forcats::fct_reorder(INJ_SEVNAME, injury_share, .fun = max, .desc = TRUE))

ggplot(injury_severity_by_vehicle_type, aes(x = INJ_SEVNAME, y = injury_share, fill = vehicle_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Injury Severity Mix by Vehicle Type",
    x = "Injury Severity",
    y = "Proportion within Vehicle Type",
    fill = "Vehicle Type"
  ) +
  theme_minimal()
```

This grouped bar chart shifts attention from crash-level fatalities to the distribution of injury severity categories within each vehicle type. According to Brockhus et al. (2024), injury severity in motorcycle crashes is strongly associated with multiple interacting crash conditions, so a visibly different severity mix here would be consistent with that broader descriptive literature. It also raises a follow-up question about whether ejection patterns differ across motorcycles and cars in a way that mirrors the severity distribution.

```{r bar-ejection-by-vehicle-type, echo=TRUE}
ejection_rate_by_vehicle_type <- fars_full %>%
  filter(!is.na(vehicle_type), !is.na(EJECTIONNAME)) %>%
  mutate(
    ejected = case_when(
      str_detect(as.character(EJECTIONNAME), regex("not ejected", ignore_case = TRUE)) ~ "Not Ejected",
      TRUE ~ "Any Ejection"
    )
  ) %>%
  count(vehicle_type, ejected) %>%
  group_by(vehicle_type) %>%
  mutate(ejection_share = n / sum(n)) %>%
  ungroup()

ggplot(ejection_rate_by_vehicle_type, aes(x = vehicle_type, y = ejection_share, fill = ejected)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Ejection Pattern by Vehicle Type",
    x = "Vehicle Type",
    y = "Proportion of Records",
    fill = "Ejection Status"
  ) +
  theme_minimal()
```

Ejection is a useful descriptive bridge between vehicle type and injury outcomes because the share of records involving any ejection can be compared directly across the two groups. A noticeably higher ejection share for motorcycles would fit the expectation that motorcycles expose riders to different crash environments than enclosed passenger cars. That, in turn, makes it useful to ask whether the motorcycle-car contrast is stable over time or concentrated in particular years.

```{r line-fatality-trend, echo=TRUE}
fatality_trend_by_year <- crash_vehicle_records %>%
  filter(!is.na(YEAR), !is.na(vehicle_type), !is.na(FATALS)) %>%
  group_by(YEAR, vehicle_type) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop")

ggplot(fatality_trend_by_year, aes(x = YEAR, y = total_fatalities, color = vehicle_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2011:2020) +
  labs(
    title = "Fatality Counts over Time by Vehicle Type",
    x = "Year",
    y = "Fatality Count",
    color = "Vehicle Type"
  ) +
  theme_minimal()
```

The year-by-year line chart shows whether the motorcycle-car gap is persistent across the full 2011–2020 span rather than driven by a single outlier year. According to IIHS (2024), the overrepresentation of motorcyclists in traffic deaths is not a one-year anomaly, so a consistently elevated motorcycle series would reinforce that longer-run pattern. Once the broad time trend is established, the next set of plots narrows the focus to motorcycle-only records and asks how helmet use varies across place and outcomes.

```{r bar-helmet-rate-by-state, echo=TRUE}
helmet_rate_by_state <- motorcycle_records %>%
  filter(!is.na(STATENAME), !is.na(helmet_used), STATENAME %in% top_motorcycle_states) %>%
  group_by(STATENAME) %>%
  summarise(helmet_rate = mean(helmet_used), .groups = "drop") %>%
  mutate(STATENAME = forcats::fct_reorder(STATENAME, helmet_rate))

ggplot(helmet_rate_by_state, aes(x = STATENAME, y = helmet_rate)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Helmet Use Rate by State among Motorcycle Records",
    x = "State",
    y = "Helmet Use Rate"
  ) +
  theme_minimal()
```

This state comparison looks only at motorcycle records and summarizes how often helmet use is recorded among the highest-count states in the sample. Because `STATENAME` is available only for the 2020 portion of this pipeline, the figure should be read as a one-year geographic snapshot rather than a decade-long state comparison. The spread across states naturally leads to the more direct question of how helmet use is associated with fatal outcomes within motorcycle crashes.

```{r bar-fatality-by-helmet, echo=TRUE}
fatality_by_helmet_use <- motorcycle_records %>%
  filter(!is.na(helmet_used), !is.na(FATALS)) %>%
  group_by(helmet_used) %>%
  summarise(fatality_rate = mean(FATALS > 0), .groups = "drop") %>%
  mutate(helmet_label = if_else(helmet_used, "Helmet Used", "No Helmet Recorded"))

ggplot(fatality_by_helmet_use, aes(x = helmet_label, y = fatality_rate, fill = helmet_label)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Fatality Rate by Helmet Use among Motorcycle Records",
    x = "Helmet Use",
    y = "Proportion of Records with FATALS > 0"
  ) +
  theme_minimal()
```

This is one of the clearest motorcycle-only comparisons because it puts helmet use next to the share of records that involve at least one fatality. According to NHTSA (2024), helmet use is one of the most effective interventions for reducing motorcyclist fatality risk, so any visible gap between these two bars would align with that descriptive safety evidence. After comparing fatality rates, it makes sense to examine whether the full injury severity profile also differs by helmet status.

```{r bar-inj-sev-by-helmet-faceted, echo=TRUE}
injury_severity_by_helmet <- motorcycle_records %>%
  filter(!is.na(helmet_used), !is.na(INJ_SEVNAME)) %>%
  mutate(helmet_label = if_else(helmet_used, "Helmet Used", "No Helmet Recorded")) %>%
  count(helmet_label, INJ_SEVNAME) %>%
  group_by(helmet_label) %>%
  mutate(injury_share = n / sum(n)) %>%
  ungroup() %>%
  mutate(INJ_SEVNAME = forcats::fct_reorder(INJ_SEVNAME, injury_share, .fun = max, .desc = TRUE))

ggplot(injury_severity_by_helmet, aes(x = INJ_SEVNAME, y = injury_share, fill = helmet_label)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ helmet_label) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Injury Severity Distribution by Helmet Use among Motorcycle Records",
    x = "Injury Severity",
    y = "Proportion within Helmet Group"
  ) +
  theme_minimal()
```

Faceting keeps the injury-severity comparison readable while still preserving the full category mix for helmeted and non-helmeted motorcycle records. Brockhus et al. (2024) emphasize that serious motorcycle injuries reflect multiple interacting conditions, so differences between these two panels should be interpreted as descriptive associations rather than a single-factor explanation. The next step is to move outward again and ask whether environmental conditions such as weather and lighting also line up with motorcycle fatalities.

```{r bar-fatality-by-weather, echo=TRUE}
top_motorcycle_weather <- motorcycle_crash_records %>%
  filter(!is.na(WEATHERNAME)) %>%
  count(WEATHERNAME, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(WEATHERNAME) %>%
  as.character()

motorcycle_fatalities_by_weather <- motorcycle_crash_records %>%
  filter(!is.na(WEATHERNAME), !is.na(FATALS), WEATHERNAME %in% top_motorcycle_weather) %>%
  group_by(WEATHERNAME) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop") %>%
  mutate(WEATHERNAME = forcats::fct_reorder(WEATHERNAME, total_fatalities))

ggplot(motorcycle_fatalities_by_weather, aes(x = WEATHERNAME, y = total_fatalities)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Motorcycle Fatalities by Weather Condition",
    x = "Weather Condition",
    y = "Fatality Count"
  ) +
  theme_minimal()
```

Restricting the plot to the top ten weather categories keeps the comparison readable while still showing where motorcycle fatalities are concentrated across common conditions. Since `WEATHERNAME` is available only in the 2020-style descriptive columns, this figure should be treated as a limited-condition snapshot rather than a full multi-year environmental trend. That same limitation applies to the next plot on lighting, which asks whether fatalities cluster differently across daylight and dark conditions.

```{r bar-fatality-by-lgt-cond, echo=TRUE}
motorcycle_fatalities_by_light <- motorcycle_crash_records %>%
  filter(!is.na(LGT_CONDNAME), !is.na(FATALS)) %>%
  group_by(LGT_CONDNAME) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop") %>%
  mutate(LGT_CONDNAME = forcats::fct_reorder(LGT_CONDNAME, total_fatalities))

ggplot(motorcycle_fatalities_by_light, aes(x = LGT_CONDNAME, y = total_fatalities)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Motorcycle Fatalities by Light Condition",
    x = "Light Condition",
    y = "Fatality Count"
  ) +
  theme_minimal()
```

The lighting chart complements the weather comparison by asking whether motorcycle fatalities are concentrated in daylight, dark-lighted, or dark-unlighted settings. Together, the two environmental figures help identify descriptive conditions that may be worth combining with rider-level factors later, while still avoiding any causal claim. From there, the analysis returns to a broader comparison of setting and vehicle type using rural-urban categories.

```{r bar-rur-urb-by-vehicle-type, echo=TRUE}
fatality_by_setting_and_vehicle <- crash_vehicle_records %>%
  filter(!is.na(RUR_URBNAME), !is.na(vehicle_type), !is.na(FATALS)) %>%
  group_by(vehicle_type, RUR_URBNAME) %>%
  summarise(fatality_rate = mean(FATALS > 0), .groups = "drop") %>%
  mutate(RUR_URBNAME = forcats::fct_reorder(RUR_URBNAME, fatality_rate, .fun = max, .desc = TRUE))

ggplot(fatality_by_setting_and_vehicle, aes(x = RUR_URBNAME, y = fatality_rate, fill = vehicle_type)) +
  geom_col(position = "dodge") +
  facet_wrap(~ vehicle_type) +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_flip() +
  labs(
    title = "Fatality Proportions by Rural-Urban Setting and Vehicle Type",
    x = "Rural or Urban Setting",
    y = "Proportion of Crash Records with FATALS > 0",
    fill = "Vehicle Type"
  ) +
  theme_minimal()
```

This faceted setting comparison checks whether the motorcycle-car fatality gap looks similar across rural and urban contexts once missing values are removed. Because the plot organizes the same setting categories inside separate vehicle-type panels, it is easier to compare patterns without overcrowding the graphic. A final natural follow-up is whether travel speed differs across injury severity levels in a way that also varies by vehicle type.

```{r bar-mean-trav-sp-by-vehicle-inj, echo=TRUE}
mean_speed_by_vehicle_and_injury <- fars_full %>%
  filter(!is.na(vehicle_type), !is.na(INJ_SEVNAME), !is.na(TRAV_SP), INJ_SEVNAME %in% top_injury_levels) %>%
  group_by(vehicle_type, INJ_SEVNAME) %>%
  summarise(mean_travel_speed = mean(TRAV_SP), .groups = "drop") %>%
  mutate(INJ_SEVNAME = forcats::fct_reorder(INJ_SEVNAME, mean_travel_speed, .fun = max))

ggplot(mean_speed_by_vehicle_and_injury, aes(x = INJ_SEVNAME, y = mean_travel_speed, fill = vehicle_type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Mean Travel Speed by Vehicle Type and Injury Severity",
    x = "Injury Severity (Top 4 Levels)",
    y = "Mean Travel Speed (mph)",
    fill = "Vehicle Type"
  ) +
  theme_minimal()
```

Ending with mean travel speed adds one more layer of complexity by comparing a continuous variable across both vehicle type and injury severity categories. This figure is still descriptive, but it helps show whether the severity patterns explored earlier are accompanied by different typical speed levels within the top injury categories. That closes the progression from simple two-variable summaries to more structured comparisons involving time, place, environment, and rider protection.

### Exploration Log

I started this section with the broadest possible comparison: motorcycles versus cars on fatality counts and fatal-crash proportions. That first pass was useful because it established the main descriptive gap quickly, but it also made clear that raw totals alone were too blunt to explain much, so I moved next to the faceted fatality histogram and the injury-severity comparison. Those figures made the section feel more coherent because they shifted the story from “are the totals different?” to “how do the distributions of harm differ across the two vehicle groups?”

One surprise was how naturally the ejection comparison fit into the sequence. I originally thought of it as a side plot, but it turned into a helpful bridge between the vehicle-type comparisons and the motorcycle-only helmet analysis because it pointed toward exposure and rider vulnerability without forcing a causal interpretation. A smaller dead end was the temptation to stack too many categories into a single multicolor bar chart; moving to faceting for the more complex figures made the progression much easier to read and kept the section aligned with the project guardrails.

From there, one question led to the next in a fairly direct way. Once the motorcycle-car gap looked persistent over time, it made sense to zoom into motorcycle-only records and ask how helmet use differed across states and how helmet status was associated with fatality and injury severity. After that, the environmental plots on weather and light conditions added context, and the rural-urban and travel-speed figures closed the section by showing how broader crash setting variables can be layered onto the earlier relationships. Overall, the exploration moved from a simple vehicle comparison into a more connected descriptive narrative about severity, protection, setting, and conditions.

## Section C — Five Poster Figures

### Poster Figure 1

```{r poster-fig-1, echo=TRUE}
poster_fig_1 <- crash_vehicle_records %>%
  filter(!is.na(YEAR), !is.na(vehicle_type), !is.na(FATALS)) %>%
  group_by(YEAR, vehicle_type) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop")

ggplot(poster_fig_1, aes(x = YEAR, y = total_fatalities, color = vehicle_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = 2011:2020) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Motorcycle vs. Car Fatalities, 2011–2020",
    x = "Year",
    y = "Total Fatalities",
    color = "Vehicle Type"
  ) +
  theme_minimal(base_size = 14)
```

This setup figure opens the poster with the clearest broad comparison in the project: how fatalities linked to motorcycle and car crash records move over the full 2011–2020 window. It earns the first spot in the narrative because a viewer can immediately see the decade-long pattern before getting pulled into more detailed breakdowns. IIHS (2024) emphasizes that motorcyclists are overrepresented in traffic deaths relative to their presence on the road, so this figure anchors the poster in that larger descriptive context. The takeaway is that the motorcycle-versus-car contrast is not a one-panel curiosity but the starting point for the rest of the story.

### Poster Figure 2

```{r poster-fig-2, echo=TRUE}
poster_fig_2 <- fars_full %>%
  filter(!is.na(vehicle_type), !is.na(INJ_SEVNAME)) %>%
  mutate(injury_group = forcats::fct_lump_prop(as.factor(INJ_SEVNAME), prop = 0.03, other_level = "Other")) %>%
  count(vehicle_type, injury_group) %>%
  group_by(vehicle_type) %>%
  mutate(injury_share = n / sum(n)) %>%
  ungroup() %>%
  mutate(injury_group = forcats::fct_reorder(injury_group, injury_share, .fun = max, .desc = TRUE))

ggplot(poster_fig_2, aes(x = injury_group, y = injury_share, fill = vehicle_type)) +
  geom_col(position = position_dodge(width = 0.8)) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Injury Severity Distribution: Motorcycles vs. Cars",
    x = "Injury Severity",
    y = "Proportion within Vehicle Type",
    fill = "Vehicle Type"
  ) +
  theme_minimal(base_size = 14)
```

This zoom-in figure takes the viewer from the broad fatality trend into the composition of injury outcomes within each vehicle group. Collapsing rarer severity categories keeps the poster readable while still preserving the main differences in the shape of the distributions. It belongs second in the arc because it answers the natural follow-up to Figure 1: not just whether motorcycles and cars differ, but where that difference appears in observed injury severity. A viewer should come away seeing that the overall contrast also shows up inside the injury mix, not only in total counts.

### Poster Figure 3

```{r poster-fig-3, echo=TRUE}
poster_fig_3 <- motorcycle_records %>%
  filter(
    YEAR == 2020,
    !is.na(STATENAME),
    !is.na(helmet_used),
    STATENAME %in% top_motorcycle_states
  ) %>%
  group_by(STATENAME) %>%
  summarise(helmet_rate = mean(helmet_used), .groups = "drop") %>%
  mutate(STATENAME = forcats::fct_reorder(STATENAME, helmet_rate))

ggplot(poster_fig_3, aes(x = STATENAME, y = helmet_rate, fill = helmet_rate)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_gradient(low = "tomato", high = "steelblue") +
  labs(
    title = "Helmet Use Rate by State (Top 15 Motorcycle States, 2020)",
    x = "State",
    y = "Helmet Use Rate",
    fill = "Helmet Rate"
  ) +
  theme_minimal(base_size = 14)
```

This nuance figure adds geographic texture by showing that helmet use among motorcycle records is not uniform across the highest-count states. It sits in the middle of the poster because it complicates the earlier comparison without abandoning the clear helmet-safety thread that the audience is already following. Since `STATENAME` is populated only for the 2020 data in this pipeline, the figure should be read as a 2020 snapshot rather than a decade-long state ranking. The key takeaway is that motorcycle safety patterns vary across place, so the national story has meaningful internal variation.

### Poster Figure 4

```{r poster-fig-4, echo=TRUE}
poster_fig_4 <- motorcycle_crash_records %>%
  filter(!is.na(helmet_used), !is.na(FATALS)) %>%
  group_by(helmet_used) %>%
  summarise(fatality_rate = mean(FATALS > 0), .groups = "drop")

ggplot(poster_fig_4, aes(x = as.character(helmet_used), y = fatality_rate, fill = as.character(helmet_used))) +
  geom_col(width = 0.65) +
  scale_x_discrete(labels = c("TRUE" = "Helmet Used", "FALSE" = "No Helmet Recorded")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Helmet Used", "FALSE" = "No Helmet Recorded")
  ) +
  labs(
    title = "Fatality Rate by Helmet Use (Motorcycle Crashes Only)",
    x = "Helmet Use",
    y = "Proportion of Crash Records with FATALS > 0",
    fill = "Helmet Use"
  ) +
  theme_minimal(base_size = 14)
```

This surprise figure reduces the poster to the most direct comparison in the whole sequence: two bars and one stark descriptive gap. That simplicity makes it ideal for a poster because viewers can interpret it quickly and carry the comparison with them as they move to the final panel. NHTSA (2024) identifies helmets as a major protective factor for motorcyclists, and this figure visually echoes that broader safety evidence without claiming that helmet use alone explains every difference in the data. The intended takeaway is a strong descriptive association between recorded helmet use and lower fatal involvement in motorcycle crashes.

### Poster Figure 5

```{r poster-fig-5, echo=TRUE}
poster_fig_5 <- motorcycle_records %>%
  filter(!is.na(helmet_used), !is.na(RUR_URBNAME), !is.na(INJ_SEVNAME)) %>%
  mutate(
    helmet_used = factor(helmet_used, levels = c(TRUE, FALSE)),
    INJ_SEVNAME = forcats::fct_lump_prop(as.factor(INJ_SEVNAME), prop = 0.03, other_level = "Other")
  ) %>%
  count(helmet_used, RUR_URBNAME, INJ_SEVNAME)

ggplot(poster_fig_5, aes(x = INJ_SEVNAME, y = n, fill = RUR_URBNAME)) +
  geom_col(position = "fill") +
  coord_flip() +
  facet_wrap(
    ~ helmet_used,
    labeller = labeller(helmet_used = c(`TRUE` = "Helmet Used", `FALSE` = "No Helmet Recorded"))
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Injury Severity by Helmet Use and Rural/Urban Setting (Motorcycle Crashes)",
    x = "Injury Severity",
    y = "Proportion within Injury Severity",
    fill = "Rural/Urban Setting"
  ) +
  theme_minimal(base_size = 14)
```

This big-picture figure closes the arc by combining rider protection, injury severity, and crash setting into one structured view. Faceting by helmet status keeps the comparison readable, while the fill colors show how rural and urban settings are distributed within each severity category. Because `RUR_URBNAME` is only populated in the 2020 descriptive fields used here, the panel works best as a contextual snapshot rather than a full-period summary. A viewer should leave with the sense that motorcycle injury patterns sit inside a broader environment of protection choices and crash settings, not just a single headline difference.

## Section D — Variable Rationale

| Variable Name | Source Dataset | Type | Rationale for Inclusion |
|---|---|---|---|
| `ST_CASE` | accident / vehicle / person | Numeric (ID) | This crash-case identifier is the backbone of the full join because it lets the project connect crash, vehicle, and person records into one analysis file. Without it, none of the multi-table comparisons in the EDA would be possible. |
| `YEAR` | accident | Numeric (Discrete) | `YEAR` makes the full 2011–2020 window usable for trend analysis and keeps crash records from different years from being merged incorrectly. It is especially important for poster figures that compare patterns across the decade rather than in a single snapshot. |
| `STATENAME` | accident | Character | `STATENAME` provides the geographic context for the state-level helmet comparison in the poster section. Because it is only populated in the 2020 descriptive fields used here, it supports a one-year location snapshot rather than a full-period state trend. |
| `FATALS` | accident | Numeric (Discrete) | `FATALS` is the central crash-level outcome in the document because it records how many people died in each crash. It drives the broadest comparisons in the EDA, including total fatalities, fatal-crash proportions, and several poster figures. |
| `DRUNK_DR` | accident | Numeric (Discrete) | The count of drunk drivers in a crash adds one of the most policy-relevant behavioral risk factors in the file. Even when it is not always centered in the final poster story, it remains important for describing the severity context of motorcycle and car crashes. |
| `RUR_URBNAME` | accident | Character | This variable captures whether crashes occurred in rural or urban settings, which helps extend the analysis beyond rider behavior into crash environment. In the EDA, it supports the final poster figure and the broader question of whether motorcycle injury patterns differ across settings. |
| `FUNC_SYSNAME` | accident | Character | `FUNC_SYSNAME` identifies the roadway class associated with each crash, such as interstate or local roads. That makes it useful for describing whether motorcycle risk patterns appear in different roadway environments, even though those descriptive labels are only available for 2020 in this pipeline. |
| `WEATHERNAME` | accident | Character | Weather conditions matter because motorcycles are more exposed to rain, fog, and other adverse environments than enclosed passenger vehicles. This variable supports the EDA's environmental-condition plots and helps interpret whether fatalities are concentrated in common or adverse weather categories. |
| `LGT_CONDNAME` | accident | Character | Lighting conditions provide a second environmental lens by distinguishing daylight from dark-lighted and dark-unlighted settings. It is included because visibility is a plausible descriptive factor in motorcycle safety patterns and because it supports one of the focused environmental plots in Section B. |
| `MAN_COLLNAME` | accident | Character | The manner-of-collision field adds information about crash configuration, such as angle or rear-end impacts. Including it broadens the EDA beyond simple outcome counts and helps characterize the kinds of crash events represented in the joined file. |
| `DAY_WEEKNAME` | accident | Character | Day of week is a compact timing variable that helps summarize whether crashes are spread evenly across the week or concentrated on particular days. It gives the EDA a basic temporal rhythm beyond the year trend and can help frame behavioral or travel-pattern differences. |
| `HOURNAME` | accident | Character | `HOURNAME` refines the timing story by shifting from day-level to time-of-day patterns. It is useful because motorcycle exposure and visibility conditions can vary sharply across hours, even if the final poster emphasizes broader trends and helmet comparisons. |
| `BODY_TYPNAME` | vehicle | Character | `BODY_TYPNAME` is the clearest descriptive vehicle classification in the source files and is the main ingredient for distinguishing motorcycles from passenger cars in later years. Because body-type names are only consistently available from 2015–2020, they also highlight the schema drift the EDA has to manage. |
| `ROLLOVERNAME` | vehicle | Character | Rollover is a meaningful crash-dynamics variable because it captures whether the crash involved vehicle overturning. Even if it is not one of the headline poster variables, it contributes to the broader severity and crash-mechanism context in the EDA. |
| `TRAV_SP` | vehicle | Numeric (Continuous) | Travel speed is one of the most interpretable severity-related variables in the vehicle file. It appears in the relationship section because comparing average speeds across injury categories helps connect crash outcomes to a continuous measure of crash intensity. |
| `DEFORMEDNAME` | vehicle | Character | Reported vehicle deformation offers a descriptive proxy for crash damage severity. Including it helps the EDA summarize how severe the physical impact appears across records, which complements the direct fatality and injury measures. |
| `L_STATUSNAME` | vehicle | Character | License status provides another rider or driver background variable that can matter for interpreting crash risk patterns. It is included because legal driving status is substantively relevant to safety discussions even when it is not a centerpiece of the final poster arc. |
| `AGE` | person | Numeric (Continuous) | Age is one of the simplest and most necessary person-level descriptors in the file. The EDA uses it to summarize who appears in the joined records and to keep the reader grounded in the demographic composition of the analysis sample. |
| `SEXNAME` | person | Character | `SEXNAME` adds another core demographic dimension for describing the person-level records. It helps show whether the sample is heavily concentrated in one group and provides context for later severity and helmet-use comparisons. |
| `PER_TYPNAME` | person | Character | Person type distinguishes drivers, passengers, and other roles, which matters when interpreting who the person-level records actually represent. It supports the project's effort to focus on motorcycle operators and comparable occupants rather than treating all people in the files as interchangeable. |
| `INJ_SEVNAME` | person | Character | Injury severity is the main person-level outcome in the EDA and one of the most informative complements to the crash-level fatality count. It supports several of the strongest comparisons in the document, especially the vehicle-type and helmet-based severity distributions. |
| `REST_USENAME` | person | Character | `REST_USENAME` is the main helmet-related source field for the earlier years in the pipeline, especially 2011–2015. It is included because the project needs a consistent helmet indicator over time, and this variable supplies much of the information used to build that measure. |
| `EJECTIONNAME` | person | Character | Ejection status is a strong descriptive bridge between vehicle type and injury severity because it captures whether a person was thrown from the vehicle. It helps the EDA tell a more physical story about exposure differences between motorcycles and cars. |
| `DRINKINGNAME` | person | Character | This person-level drinking indicator adds another behavioral safety dimension beyond the crash-level drunk-driver count. It is useful for describing whether alcohol involvement appears frequently in the records and for framing severity patterns in a broader risk context. |
| `vehicle_type` | derived from `BODY_TYPNAME` and `BODY_TYP` | Factor (Derived Binary Category) | `vehicle_type` is a key derived variable because the entire project depends on a clean motorcycle-versus-car comparison across years with inconsistent schemas. Constructing it from both named and numeric body-type fields lets the EDA preserve the full 2011–2020 window instead of restricting analysis to the years with descriptive labels. |
| `helmet_used` | derived from `REST_USENAME` and `HELM_USENAME` | Logical (Derived Binary Indicator) | `helmet_used` is the project's main rider-protection variable and one of the central inputs to the poster story. It combines `REST_USENAME` for earlier years with `HELM_USENAME` where available so that helmet-related comparisons are possible across the decade rather than only in the later files. |

### Expert Sources for EDA Interpretation

1. **National Highway Traffic Safety Administration. _Traffic Safety Facts: Motorcycles — 2023 Data_. U.S. Department of Transportation, 2024.** <https://crashstats.nhtsa.dot.gov/Api/Public/ViewPublication/813732.pdf>

   This federal report provides the clearest national benchmark for interpreting the fatality and helmet-use patterns shown in the EDA. It is especially helpful for Figures 1 and 4 because it explains why helmet use and fatality differences are central, not peripheral, to motorcycle safety analysis. In this document, I use it less as a source of causal claims and more as a credibility anchor for the descriptive comparisons in the poster figures. It helps show that the patterns in the FARS-based EDA fit into a broader national safety conversation rather than standing alone as isolated classroom results.

2. **Insurance Institute for Highway Safety. _Fatality Facts: Motorcycles and ATVs_. IIHS, 2024.** <https://www.iihs.org/research-areas/fatality-statistics/detail/motorcycles-and-atvs>

   The IIHS summary is useful because it translates national motorcycle fatality patterns into a compact, readable source that supports the broad vehicle-type comparisons in the EDA. It is particularly relevant for the opening fatality figures because those plots ask viewers to compare motorcycles and cars at a high level before moving into more detailed breakdowns. For the EDA, this source helps justify why a motorcycle-versus-car contrast is a meaningful starting point for the poster narrative. It also gives outside support for describing motorcyclists as disproportionately represented in traffic fatalities.

3. **Brockhus, L. A., et al. “Injury Patterns and Outcomes in Motorcycle Driver Crashes.” _Accident Analysis & Prevention_, Elsevier, 2024.**

   This article is most helpful for the injury-severity and ejection portions of the EDA because it emphasizes that motorcycle crash outcomes reflect multiple interacting conditions rather than one isolated factor. That perspective fits the document's progression from simple fatality comparisons to richer views involving severity, ejection, helmet use, and environment. In the EDA context, the study supports treating the plotted relationships as descriptive patterns that should be interpreted alongside crash setting and rider protection variables. It strengthens the decision to include several severity-related variables in the rationale table instead of centering only on fatalities.

4. **Doerzaph, Z., et al. _Injury Risk Curves for Motorcycles in the United States_. Virginia Tech Transportation Institute, 2025.**

   This source is especially relevant for interpreting variables tied to crash intensity and context, such as travel speed, roadway setting, and collision conditions. In the EDA, it helps explain why speed and environment deserve a place in the poster even when they are not the single headline finding. The study also supports the idea that motorcycle injury risk should be read through combinations of conditions rather than through a one-variable story. That makes it a useful backdrop for the more layered figures in Section B and for the rationale behind keeping `TRAV_SP`, `RUR_URBNAME`, and roadway-condition variables in the analysis file.

5. **Wang, M. H., et al. “Factors Contributing to Motorcyclist Fatalities.” _Transportation Safety Journal_, 2022.** <https://pmc.ncbi.nlm.nih.gov/articles/PMC9318472>

   Wang et al. are helpful because they focus directly on the kinds of crash environments and rider characteristics that appear throughout this EDA. Their work supports the choice to include variables such as roadway setting, lighting, weather, and person-level characteristics even when some of those fields are only fully populated in 2020. For the EDA, the article reinforces that motorcycle fatality patterns are shaped by a combination of environment, crash structure, and rider factors. That framing is valuable when explaining why the poster does not stop at a single helmet comparison but instead builds a broader descriptive story around severity and setting.

## Section E — Self-Assessment

**1. Which poster figures do I feel most confident about, and why?**

I feel most confident about Figures 1 and 4. Figure 1 uses `YEAR` across the full 2011–2020 period, so it rests on one of the most complete and stable dimensions in the cleaned file and avoids the narrower 2020-only descriptive fields. Figure 4 is also strong because it asks a very direct question with a clean binary comparison: among motorcycle crash records, how does the fatal-crash share differ by `helmet_used`? Together, those two figures are the easiest to defend because they rely on broadly available variables, simple visual structure, and a story that matches the main purpose of the project.

**2. Which poster figures do I feel least confident about, and why?**

I feel least confident about Figures 3 and 5. Both depend on variables that are only populated in the 2020 descriptive data used here, especially `STATENAME` in Figure 3 and `RUR_URBNAME` in Figure 5, so they are better understood as snapshots than as decade-wide summaries. That limitation does not make the figures useless, but it does narrow what I can claim from them and makes them more sensitive to how clearly I explain their scope. Compared with Figures 1 and 4, these two require more caveats, which makes them harder to present as central evidence.

**3. Describe one dead end in the EDA and what I learned from it.**

One dead end was my initial impulse to build per-mile fatality rates so the motorcycle-versus-car comparison would be normalized by exposure. That would have been appealing because it sounds more precise than raw counts or per-crash proportions, but FARS by itself does not provide the vehicle miles traveled denominator I would need at the same level as the crash records in this file. Once that became clear, I learned that the more defensible move was to scope the analysis to quantities I could actually support from the dataset, especially per-crash proportions and within-sample comparisons. That decision made the EDA narrower, but it also made the final figures more honest and easier to explain.

**4. What feedback would be most helpful before turning this into the final poster?**

The most helpful feedback would be on story-arc clarity for readers who do not already know how FARS works. I would especially like input on whether the poster makes the transition from broad vehicle comparisons to helmet and setting comparisons feel natural, or whether that sequence still needs stronger explanation. I would also want feedback on the defensibility of how I constructed `helmet_used`, since it combines `REST_USENAME` and `HELM_USENAME` across years with slightly different schemas. If those two points are clear and persuasive, I think the poster will be much easier for an outside reader to trust.
